New cardinality estimation algorithms for HyperLogLog sketches

### 2016-07-17 paper/paper.tex
some minor changes
minor changes

\title{New cardinality estimation algorithms for HyperLogLog sketches}
\subtitle{Draft version}
\author{Otmar Ertl \\ otmar.ertl@gmail.com}
\begin{document}
\maketitle
\let\thefootnote\relax\footnotetext{This article together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}.}
\begin{abstract}
This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches. A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities. Based on the maximum likelihood principle another unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate. The maximum likelihood method is also appropriate to improve cardinality estimates of set intersections compared to the inclusion-exclusion principle. The new methods are demonstrated and verified by extensive simulations.
\end{abstract}

@@

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean (HM) was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw estimator which is given by
\begin{equation}
\label{equ:raw_estimator}
\symCardinalityRawEstimate
=
2\alpha_\symNumReg
\operatorname{HM}
\!\left(
\symNumReg 2^{\symRegValVariate_1-1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg-1}
\right)
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexI=1}^{\symNumReg}2^{-\symRegValVariate_\symIndexI}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCountVariate_\symRegVal 2^{-\symRegVal}}.
\end{equation}
Here $\alpha_\symNumReg$ is a bias correction factor which was derived for a given number of registers $\symNumReg$ to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\!\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{\!\!-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision$. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:raw_estimator}, because the additional bias is negligible compared to the estimation error.

@@

\item If $\symX$ is smaller than 0.5, the function $\symHelper(\symX)$ can be well approximated by a Taylor series around $\symX=0$
\begin{equation}
\symHelper(\symX)
=
\frac{\symX}{2} - \frac{\symX^2}{12} +\frac{\symX^4}{720}-\frac{\symX^6}{30240} + \symBigO(\symX^{8})
\end{equation}
which can be optimized for numerical evaluation using Estrin's scheme and $\symX' := \frac{\symX}{2}$ and $\symX'' := \symX' \symX'$
\begin{equation}
\label{equ:taylor}
\symHelper(\symX)
=
\symX' - \symX''/3 + \left(\symX'' \symX''\right)\left(1/45-\symX''/472.5\right)
+ \symBigO(\symX^{8})
\end{equation}
The smallest argument for which $\symHelper$ needs to be evaluated is $\frac{\symX}{2^{\symRegValVariate'_\text{max}}}$. If 
$\symCountVariate_{\symRegRange+1}=0$, $\frac{\symX}{2^{\symRegValVariate'_\text{max}}} \leq \log 2 \approx 0.693$ holds, which is straightforward to see using \eqref{equ:strong_upper_bound}. In practice, $\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\leq0.5$ is satisfied most of the time as long as only a few registers are saturated ($\symCountVariate_{\symRegRange+1}\ll \symNumReg$). In case $\frac{\symX}{2^{\symRegValVariate'_\text{max}}} > 0.5$, $\symHelper\!\left(\frac{\symX}{2^{\symKappa}}\right)$ is calculated instead with $\symKappa = 2+\lfloor\log_2(\symX)\rfloor$. By definition, $\frac{\symX}{2^{\symKappa}}\leq 0.5$ which allows using the Taylor series approximation. $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ is finally obtained after $\symKappa -\symRegValVariate'_\text{max}$ iterations using \eqref{equ:helper_recursion1}.
\end{itemize}

@@

The HyperLogLog sketch that represents $\symSetS_1$ can be thought to be constructed from two HyperLogLog sketches representing $\symSetA$ and $\symSetX$ and merging both according to \eqref{equ:hll_union}. Analogously, the HyperLogLog sketch for $\symSetS_2$ could have been obtained from sketches for $\symSetB$ and $\symSetX$. Let  $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ be the value of the considered register in the HyperLogLog sketch representing $\symSetA$, $\symSetB$, and $\symSetX$, respectively. The corresponding values in sketches for $\symSetS_1$ and $\symSetS_2$ are given by
\begin{equation}
\symRegValVariate_1 = \max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right)
,\quad
\symRegValVariate_2 = \max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right).
\end{equation}
Their joint cumulative probability function is given as
\begin{align}
\symProbability\!\left(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
\right)
&=
\symProbability\!\left(
\max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_1
\wedge
\max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_2
\right)
\nonumber\\
&=
\symProbability
\!\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\wedge
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\wedge
\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
\right)
\nonumber\\
&=
\symProbability
\!\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\right)
\symProbability
\!\left(
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\right)
\symProbability
\!\left(
\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
\right)
\end{align}
Here the last transformation used the independence of $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$, because by definition, the sets $\symSetA$, $\symSetB$, and $\symSetX$ are disjoint. Furthermore, under the Poisson model $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ obey 
\eqref{equ:register_value_distribution}. If we take into account that elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetXSuffix$, $\symPoissonRate_\symSetASuffix$, and $\symPoissonRate_\symSetBSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
\begin{equation}
\symProbability(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
)
=\begin{cases}
0 & \symRegVal_1 < 0 \vee \symRegVal_2 < 0
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix}}{\symNumReg 2^{\symRegVal_1}}
-
\frac{\symPoissonRate_{\symSetBSuffix}}{\symNumReg 2^{\symRegVal_2}}
-
\frac{\symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\min(\symRegVal_1, \symRegVal_2)}}
}
& 0\leq\symRegVal_1 \leq \symRegRange \wedge 0\leq\symRegVal_2\leq\symRegRange
\\
e^{
-
\frac{\symPoissonRate_{\symSetBSuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_2}}
}
& 0\leq\symRegVal_2 \leq \symRegRange < \symRegVal_1
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_1}}
}
&  0\leq\symRegVal_1 \leq \symRegRange < \symRegVal_2
\\
1
&
\symRegRange < \symRegVal_1 \wedge \symRegRange < \symRegVal_2
\end{cases}
\end{equation}

@@

@@

The logarithm of the joint probability mass function can be written using Iverson bracket notation ($\left[\text{true}\right]:=1$, $\left[\text{false}\right]:=0$) as
\begin{align}
\label{equ:joint_log_pmf_single_register}
\log(\symProbabilityMass(\symRegVal_1,\symRegVal_2))
=&
-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\symRegVal_1}}
\left[\symRegVal_1\leq\symRegRange\right]
-
\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\symRegVal_2}}
\left[\symRegVal_2\leq\symRegRange\right]
-
\frac{\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1,\symRegVal_2\right)}}
\left[\symRegVal_1\leq\symRegRange\vee\symRegVal_2\leq\symRegRange\right]
\nonumber\\
&
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_1}}}\right)
\left[1\leq\symRegVal_1<\symRegVal_2\right]
\nonumber\\
&
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}\right)
\left[\symRegVal_2<\symRegVal_1\right]
\nonumber\\
&
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_2}}}\right)
\left[1\leq\symRegVal_2<\symRegVal_1\right]
\nonumber\\
&
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal_2, \symRegRange\right)}}}\right)
\left[\symRegVal_1<\symRegVal_2\right]
\nonumber\\
&
+
\log\!\left(
1
-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
+
e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
\right)
\left[1\leq\symRegVal_1=\symRegVal_2\right]
\end{align}

Since under the Poisson model, the values for different registers are independent we are now able to write the joint probability mass function for all registers in both HyperLogLog sketches
\begin{equation}
\symProbabilityMass(\boldsymbol{\symRegVal}_1,\boldsymbol{\symRegVal}_2)
=
\prod_{\symIndexI = 1}^{\symNumReg}
\prod_{\symIndexJ = 1}^{\symNumReg}
\symProbabilityMass(\symRegVal_{1\symIndexI},\symRegVal_{2\symIndexJ}).
\end{equation}

@@

\symCountVariate^{=}_{\symRegVal}

@@

where the constant $\symCountVariate^{>}_{1\symRegVal}$,

@@

\symRegValVariate_{2\symRegVal}\rbrace\right|.

@@

The log-likelihood function \eqref{equ:log_likelihood_pair} does not always have a strict global maximum point. For example, if all register values of the first HyperLogLog sketch are larger than the corresponding values in the second HyperLogLog sketch, the function can be rewritten as sum of two functions, one dependent on $\symPoissonRate_\symSetASuffix$ and the other dependent on $(\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix)$. 
The maximum is obtained, if $\symPoissonRate_\symSetASuffix = \symPoissonRateEstimate_1$ and $\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix = \symPoissonRateEstimate_2$. Here $\symPoissonRateEstimate_1$ and $\symPoissonRateEstimate_2$ are the cardinality estimates for the first and second HyperLogLog sketch, respectively. Similarly, if all register values of the first are smaller than those of the second HyperLogLog sketch, the maximum is obtained when 
$\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix = \symPoissonRateEstimate_1$ and $\symPoissonRate_\symSetBSuffix = \symPoissonRateEstimate_2$.

If we know that the two HyperLogLog sketches have been filled by elements taken from disjoint sets we can assume $\symPoissonRate_\symSetXSuffix=0$. In this case \eqref{equ:log_likelihood_pair} is separable into the sum of two log-likelihood functions that depend on $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$, respectively, and that follow \eqref{equ:log_likelihood_single}. Hence, the joint maximum likelihood estimation of $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ gives the same results as estimating them independently by maximization of \eqref{equ:log_likelihood_single}.

The first derivatives are
\begin{align}
\frac{\partial\log \symLikelihood}{\partial\symPoissonRate_\symSetASuffix}
(
\symPoissonRate_\symSetASuffix,
\symPoissonRate_\symSetBSuffix,
\symPoissonRate_\symSetXSuffix
\vert
\boldsymbol{\symRegValVariate}_1,
\boldsymbol{\symRegValVariate}_2
)
=
&
-
\frac{1}{\symNumReg}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{
  \symCountVariate^{<}_{1\symRegVal}+
  \symCountVariate^{=}_{\symRegVal}+
  \symCountVariate^{>}_{1\symRegVal}
}{2^{\symRegVal}}
+
\frac{1}{\symNumReg}
\sum_{\symRegVal=1}^{\symRegRange+1}
\frac{1}{e^{\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}-1}
\frac{\symCountVariate^{<}_{1\symRegVal}}{2^{\min\left(\symRegVal,\symRegRange\right)}}
\nonumber\\
&
+
\frac{1}{\symNumReg}\sum_{\symRegVal=1}^{\symRegRange+1}
\frac{1}{e^{\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}-1}
\frac{\symCountVariate^{>}_{1\symRegVal}}{2^{\min\left(\symRegVal,\symRegRange\right)}}
\nonumber\\
&
+
\frac{1}{\symNumReg}
\sum_{\symRegVal=1}^{\symRegRange+1}
\frac{
e^{\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-
1
}{
e^{\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-
e^{\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
-
e^{\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal,\symRegRange\right)}}}
+
1
}
\frac{\symCountVariate^{=}_{\symRegVal}
}{2^{\min\left(\symRegVal,\symRegRange\right)}}
\end{align}

>>>>

\title{New cardinality estimation algorithms for HyperLogLog sketches}
\subtitle{Draft version}
\author{Otmar Ertl \\ otmar.ertl@gmail.com}
\begin{document}
\maketitle
\let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}.}
\begin{abstract}
This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches. A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities. Based on the maximum likelihood principle another unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate. The maximum likelihood method is also appropriate to improve cardinality estimates of set intersections compared to the inclusion-exclusion principle. The new methods are demonstrated and verified by extensive simulations.
\end{abstract}

@@

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean (HM) was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw estimator which is given by
\begin{equation}
\label{equ:raw_estimator}
\symCardinalityRawEstimate
=
\alpha_\symNumReg
\operatorname{HM}
\!\left(
\symNumReg 2^{\symRegValVariate_1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg}
\right)
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexI=1}^{\symNumReg}2^{-\symRegValVariate_\symIndexI}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCountVariate_\symRegVal 2^{-\symRegVal}}.
\end{equation}
Here $\alpha_\symNumReg$ is a bias correction factor which was derived for a given number of registers $\symNumReg$ to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\!\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{\!\!-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision$. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:raw_estimator}, because the additional bias is negligible compared to the estimation error.

@@

\item If $\symX$ is smaller than 0.5, the function $\symHelper(\symX)$ can be well approximated by a Taylor series around $\symX=0$
\begin{equation}
\symHelper(\symX)
=
\frac{\symX}{2} - \frac{\symX^2}{12} +\frac{\symX^4}{720}-\frac{\symX^6}{30240} + \symBigO(\symX^{8})
\end{equation}
which can be optimized for numerical evaluation using Estrin's scheme and $\symX' := \frac{\symX}{2}$ and $\symX'' := \symX' \symX'$
\begin{equation}
\label{equ:taylor}
\symHelper(\symX)
=
\symX' - \symX''/3 + \left(\symX'' \symX''\right)\left(1/45-\symX''/472.5\right)
+ \symBigO(\symX^{8})
\end{equation}
The smallest argument for which $\symHelper$ needs to be evaluated is $\frac{\symX}{2^{\symRegValVariate'_\text{max}}}$. If 
$\symCountVariate_{\symRegRange+1}=0$, $\frac{\symX}{2^{\symRegValVariate'_\text{max}}} \leq \log 2 \approx 0.693$ holds, which is straightforward to see using \eqref{equ:strong_upper_bound}. In practice, $\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\leq0.5$ is satisfied most of the time as long as only a few registers are saturated ($\symCountVariate_{\symRegRange+1}\ll \symNumReg$). In case $\frac{\symX}{2^{\symRegValVariate'_\text{max}}} > 0.5$, $\symHelper\!\left(\frac{\symX}{2^{\symKappa}}\right)$ is calculated instead with $\symKappa = 2+\lfloor\log_2(\symX)\rfloor$. By definition, $\frac{\symX}{2^{\symKappa}}\leq 0.5$ which allows using the Taylor series approximation. $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ is finally obtained after $\symKappa -\symRegValVariate'_\text{max}$ iterations using \eqref{equ:helper_recursion1}. As shown in \cref{app:error_approx}, a small approximation error of $\symHelper$ does not have much impact on the error of the maximum likelihood estimate as long as most registers are not yet saturated.
\end{itemize}

@@

The HyperLogLog sketch that represents $\symSetS_1$ can be thought to be constructed from two HyperLogLog sketches representing $\symSetA$ and $\symSetX$ and merging both using \eqref{equ:hll_union}. Analogously, the HyperLogLog sketch for $\symSetS_2$ could have been obtained from sketches for $\symSetB$ and $\symSetX$. Let  $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ be the value of the considered register in the HyperLogLog sketch representing $\symSetA$, $\symSetB$, and $\symSetX$, respectively. The corresponding values in sketches for $\symSetS_1$ and $\symSetS_2$ are given by
\begin{equation}
\symRegValVariate_1 = \max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right)
,\quad
\symRegValVariate_2 = \max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right).
\end{equation}
Their joint cumulative probability function is given as
\begin{align}
\symProbability\!\left(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
\right)
&=
\symProbability\!\left(
\max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_1
\wedge
\max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_2
\right)
\nonumber\\
&=
\symProbability
\!\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\wedge
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\wedge
\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
\right)
\nonumber\\
&=
\symProbability
\!\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\right)
\symProbability
\!\left(
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\right)
\symProbability
\!\left(
\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
\right).
\end{align}
Here the last transformation used the independence of $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$, because by definition, the sets $\symSetA$, $\symSetB$, and $\symSetX$ are disjoint. Furthermore, under the Poisson model $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ obey 
\eqref{equ:register_value_distribution}. If we take into account that elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetXSuffix$, $\symPoissonRate_\symSetASuffix$, and $\symPoissonRate_\symSetBSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
\begin{equation}
\symProbability(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
)
=\begin{cases}
0 & \symRegVal_1 < 0 \vee \symRegVal_2 < 0
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix}}{\symNumReg 2^{\symRegVal_1}}
-
\frac{\symPoissonRate_{\symSetBSuffix}}{\symNumReg 2^{\symRegVal_2}}
-
\frac{\symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\min(\symRegVal_1, \symRegVal_2)}}
}
& 0\leq\symRegVal_1 \leq \symRegRange \wedge 0\leq\symRegVal_2\leq\symRegRange
\\
e^{
-
\frac{\symPoissonRate_{\symSetBSuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_2}}
}
& 0\leq\symRegVal_2 \leq \symRegRange < \symRegVal_1
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_1}}
}
&  0\leq\symRegVal_1 \leq \symRegRange < \symRegVal_2
\\
1
&
\symRegRange < \symRegVal_1 \wedge \symRegRange < \symRegVal_2
\end{cases}
\end{equation}

@@

),

@@

The logarithm of the joint probability mass function can be written using Iverson bracket notation ($\left[\text{true}\right]:=1$, $\left[\text{false}\right]:=0$) as
\begin{align}
\label{equ:joint_log_pmf_single_register}
\log(\symProbabilityMass(\symRegVal_1,\symRegVal_2))
=&
-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\symRegVal_1}}
\left[\symRegVal_1\leq\symRegRange\right]
-
\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\symRegVal_2}}
\left[\symRegVal_2\leq\symRegRange\right]
-
\frac{\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1,\symRegVal_2\right)}}
\left[\symRegVal_1\leq\symRegRange\vee\symRegVal_2\leq\symRegRange\right]
\nonumber\\
&
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_1}}}\right)
\left[1\leq\symRegVal_1<\symRegVal_2\right]
\nonumber\\
&
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetASuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}\right)
\left[\symRegVal_2<\symRegVal_1\right]
\nonumber\\
&
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\symRegVal_2}}}\right)
\left[1\leq\symRegVal_2<\symRegVal_1\right]
\nonumber\\
&
+
\log\!\left(1-e^{-\frac{\symPoissonRate_\symSetBSuffix}{\symNumReg 2^{\min\left(\symRegVal_2, \symRegRange\right)}}}\right)
\left[\symRegVal_1<\symRegVal_2\right]
\nonumber\\
&
+
\log\!\left(
1
-e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
+
e^{-\frac{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix}{\symNumReg 2^{\min\left(\symRegVal_1, \symRegRange\right)}}}
\right)
\left[1\leq\symRegVal_1=\symRegVal_2\right].
\end{align}

Since the values for different registers are independent under the Poisson model, we are now able to write the joint probability mass function for all registers in both HyperLogLog sketches
\begin{equation}
\symProbabilityMass(\boldsymbol{\symRegVal}_1,\boldsymbol{\symRegVal}_2)
=
\prod_{\symIndexI = 1}^{\symNumReg}
\prod_{\symIndexJ = 1}^{\symNumReg}
\symProbabilityMass(\symRegVal_{1\symIndexI},\symRegVal_{2\symIndexJ}).
\end{equation}

@@

\symCountVariate^{=}_{\symRegVal},

@@

where the constants $\symCountVariate^{>}_{1\symRegVal}$,

@@

\symRegValVariate_{2\symIndexI}\rbrace\right|.

@@

These $5\symNumReg$ values are sufficient for estimating $\symPoissonRate_\symSetA$, $\symPoissonRate_\symSetB$, and $\symPoissonRate_\symSetX$. Actually, the set of these constants can be further reduced, because $\symCountVariate^{>}_{10}=\symCountVariate^{>}_{20}=
\symCountVariate^{<}_{1(\symRegRange+1)}=\symCountVariate^{<}_{2(\symRegRange+1)}=0$
always holds.

The log-likelihood function \eqref{equ:log_likelihood_pair} does not always have a strict global maximum point. For example, if all register values of the first HyperLogLog sketch are larger than the corresponding values in the second HyperLogLog sketch, $\symCountVariate_{1\symRegVal}^{<}
=
\symCountVariate_{\symRegVal}^{=}
=
\symCountVariate_{2\symRegVal}^{>}
=
0$ for all $\symRegVal$, the function can be rewritten as sum of two functions, one dependent on $\symPoissonRate_\symSetASuffix$ and the other dependent on $\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix$. 
The maximum is obtained, if $\symPoissonRate_\symSetASuffix = \symPoissonRateEstimate_1$ and $\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix = \symPoissonRateEstimate_2$. Here $\symPoissonRateEstimate_1$ and $\symPoissonRateEstimate_2$ are the single HyperLogLog estimates for the first and second HyperLogLog sketch, respectively, which maximize the corresponding log-likelihood functions \eqref{equ:log_likelihood_single}. This means that the maximum-likelihood makes no clear statement about the intersection size in this case. The estimate for $\symPoissonRate_\symX$ could be anything between 0 and $\symPoissonRateEstimate_2$. For comparison, the inclusion-exclusion approach gives $\symPoissonRateEstimate_2$ as estimate. This result is questionable, because there is no evidence at all, that the sets $\symSetS_1$ and $\symSetS_2$ really have common elements.

If it is given that the two HyperLogLog sketches represent disjoint sets, we know that $\symPoissonRate_\symSetXSuffix=0$. In this case \eqref{equ:log_likelihood_pair} is separable into the sum of two log-likelihood functions that depend on $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$, respectively, and that follow \eqref{equ:log_likelihood_single}. Hence, the joint maximum likelihood estimation of $\symPoissonRate_\symSetASuffix$ and $\symPoissonRate_\symSetBSuffix$ gives as expected the same results as estimating them independently by maximizing \eqref{equ:log_likelihood_single}.
